Road Segment Prioritization for Bicycle Infrastructure

### 2021-03-12 README.Rmd
change methodology -> approach

---
title: Road Segment Prioritization for Bicycle Infrastructure
date: "`r Sys.Date()`"
geometry: margin=2cm
bibliography:
  - mybibfile.bib
  - rlbibfile.bib
#linenumbers: true
numbersections: true
abstract: "Understanding the motivators and deterrents to cycling is essential for creating infrastructure that gets more people to adopt cycling as a mode of transport. This paper demonstrates new methods to support the prioritisation of cycling infrastructure and cycling network design, accounting for cyclist preferences and the growing emphasis on 'filtered permeability' and 'Low Traffic Neighborhood' interventions internationally. The approach combines distance decay, route calculation, and network analysis techniques to examine where future cycling demand is most likely to arise, how such demand could be accommodated within existing street networks, and how to ensure a fair distribution of investment. Although each of these approaches have been applied to cycling infrastructure prioritisation in previous research, this is the first time that they have been combined, creating an integrated road segment prioritisation approach. The approach, which can be applied to other cities, as shown in the Appendix, is demonstrated in a case study of Manchester, resulting in cycling networks that balance directness against the need for safe and stress-free routes under different investment scenarios. A key benefit of the approach from a policy perspective is its ability to support egalitarian and cost-effective strategic cycle network planning. \\par\\textbf{Keywords:} cycling networks, low-traffic neighborhoods, routing, transport equity"
# keywords: "cycling, low-traffic neighborhoods, potential demand, routing"
output:
  # bookdown::github_document2:
  bookdown::pdf_document2:
  # # bookdown::word_document2:
      toc: false
      extra_dependencies: ["subcaption", "longtable", "float", "multirow", "placeins", "multicol"]
editor_options: 
  markdown: 
    wrap: 72
---

@@

The studies outlined above lay out the fundamentals for designing
cycling networks that generate significant cycling uptake, but they do
not propose network-level interventions. In this section we outline
techniques used in past studies, namely optimization and network
analysis techniques, such as connected components and community
detection, and examine how they are leveraged to suggest cycling network
designs. We compare the effectiveness of these network-level studies in
incorporating the fundamentals outlined above. The methodology of this
research is inspired by these techniques, but it attempts to add to them
by ensuring that all of the outlined fundamentals are accounted for. It
also goes further by attempting to factor in ethical considerations
relating to distribution of investment.

*Optimization* techniques have been used to propose improvements to
cycling networks. @mesbah2012bilevel propose a bi-level formulation to
optimize allocation of cycling lanes to the network without exceeding a
set budget. They account for the effect of cycling lanes on car traffic,
and attempt to maximize utilization of said lanes with minimal impact on
car travel times. Safeguarding against increased car traffic may be counter-productive if the goal is to create a mode-shift, as research has shown that reducing road space for cars leads to less cars on the road, a phenomenon referred to as "traffic evaporation" [@nello2020environmental].
Car usage is not considered by @mauttone2017bicycle,
who develop an optimization framework that aims to minimize the total
user cost of cycling on the network. The aggregate flow^[*flow* is
used throughout this research to refer to the cycling demand when it is
routed onto the road network. The flow on any road segment is the
cumulative demand on it, resulting from cyclists commuting between
various OD pairs] on links^[*link*
refers to a road segment throughout this research] is obtained by using
shortest paths to route existing cycling demand onto the road network,
and the solution is a proposed set of links where cycling infrastructure
should be added in order to minimize the overall travel cost of cyclists
across the network. The cost of traversing a link is given as a function
of its length and whether or not it has cycling infrastructure. The
problem has also been solved by attempting to find the minimum cost of
improving roadway links to meet a desired level of service (LOS)
[@duthie2014optimization]. In this formulation, all origin-destination
(OD) pairs need to be connected by roads that meet the desired LOS. A
limitation of these problem formulations is that they do not explicitly solve for
continuity, which is dealt with using either a constraint specifying
that each link with a bike lane should be connected to at least one
destination [@mesbah2012bilevel], a constraint on maximum deviation
from shortest paths [@duthie2014optimization], or a discontinuity
penalty to prioritize connected road segments [@mauttone2017bicycle].

@@

Our work builds on these *community finding* approaches by proposing a
similar greedy network expansion algorithm for cycle network expansion
within communities. We incorporate community finding methods for study
area partitioning with weighted routing to avoid links that are
stressful to cycle on. In doing so, we propose a methodology that
accounts for motivators and deterrents to cycling. We propose three
sub-methodologies that address on some of the limitations of previous
studies. These limitations include (a) bias inherent when basing network
design solely on existing cycling demand, (b) proposing routes that may
not correspond to studies on cyclist preference and government policies,
and (3) insufficient consideration of the ethical principles underlying
the analysis. Section \@ref(calculating-potential-cycling-demand)
focuses on calculating potential cycling demand. Section \@ref(routing)
focuses on routing the demand onto the road network while accounting for
cyclist preferences and government priorities. Section
\@ref(road-segment-prioritisation) outlines a methodology for partitioning the
study area based on a community finding algorithm and routed cycling
demand. It then introduces the network expansion algorithms, and compares an approach grounded in
'egalitarianism' to one grounded in 'utilitarianism'.

@@

\noindent where **d** and **s** are the distance and slope associated with each
OD pair and $\alpha$ and $\beta$ values are parameters calculated by a regression
model on training data. The square and square-root distance terms
"capture the non-linear impact of distance on the likelihood of
cycling", and interaction terms to capture the combined effect of slope
and distance [@lovelace2017propensity]. Alternative cycling uptake
models could be 'plugged in' to our approach for different
different contexts or scenarios of change.

@@

It should be reiterated that the weighting profile used for routing has
been developed for the purposes of this study. It creates a hierarchy of
road preference that is grounded in cyclist preferences and government
plans to create LTNs. A wide range of weighting profiles could be used
to represent different types of cyclists and road environments, as
described in @gehrke2020cycling and @furth2016network. Sensitivity
analysis could be done to determine an optimal weighting profile, but
given the variation in city road
networks<!-- (Figure \ref{fig:perc_person-km}) -->, this would probably
require calibration to the specific city. More accurate routing could be
carried out given the availability of road-level data. In such cases we
would add additional impedance to specific roads, giving more useful
routing results than the current methodology which considers all roads
of the same type to be equivalent.

@@

Given that the "most essential activity entailed in the design of
cycle-friendly infrastructure is developing a cycle network"
[@parkin2018designing], we believe that the methods have great
potential to inform investment in cities such as Manchester where there
is political will to invest in cycling long-term. A benefit of the
approach is that it has relatively modest data requirements: only the
road network (from OSM), its topography (from satellite imagery), and
OD data (from the national census in this case) are
needed, opening up the possibility of deploying the approach in other
cities^[The results are easily reproducible for all UK cities, and can
also be reproduced for cities elsewhere given the availability of
commuter data. For instructions on reproducing the results shown in this
paper, see the **README** file in the R folder of the .zip file that
accompanies this paper (to be released as open source software when the
paper is published).].

@@

The approach can identify not only where there is high
potential for cycling but also trade-offs between
stress levels (related to motor traffic) and directness.
The results can therefore be used as a basis for recommendations on
road space reallocation *and* new infrastructure to unlock potential cycling demand.
The approach encourages consideration of a wider range of preferences
and needs that previous methods that focus only on absolute potential.
Moreover, the inclusion of egalitarian principles in scenarios of change
encourages investment in cycling infrastructure to
increase the connectivity of existing cycling infrastructure
*and* investment that addresses geographical and social inequalities.
This ability to address inequalities in network prioritisation is particularly
important given research research showing substantial *inequalities*
around transport provision in general and cycling uptake and investment
in particular [@lucas2016method; @vidaltortosa_infrastructure_2020].

The approach is not without limitations. The level of detail is only as
good as the granularity of the available data (in this case relatively
coarse zones). @iacono2010measuring note that such large travel zones
are not ideal for understanding route choice behaviour of cyclists and
pedestrians. They also give rise to an \`ecological fallacy' whereby
average characteristics are assumed to apply to all residents of the
aggregated geographical area, suggesting a need for applying the methods
to more granular OD data (and for governments and other
data-collecting organisations to make OD data more
readily available).
The potential demand calculation is also based on cycling in the traditional sense, and does not consider the 
effect of micro-mobility on reducing topology-related impedance to cycling. Given that we are proposing a methodological framework which can accommodate any cycling uptake functions, this is an acceptable limitation.

>>>>

---
title: Road Segment Prioritization for Bicycle Infrastructure
date: "`r Sys.Date()`"
geometry: margin=2cm
bibliography:
  - mybibfile.bib
  - rlbibfile.bib
#linenumbers: true
numbersections: true
abstract: "Understanding the motivators and deterrents to cycling is essential for creating infrastructure that gets more people to adopt cycling as a mode of transport. This paper demonstrates a new approach to support the prioritisation of cycling infrastructure and cycling network design, accounting for cyclist preferences and the growing emphasis on 'filtered permeability' and 'Low Traffic Neighborhood' interventions internationally. The approach combines distance decay, route calculation, and network analysis methods to examine where future cycling demand is most likely to arise, how such demand could be accommodated within existing street networks, and how to ensure a fair distribution of investment. Although each of these methods have been applied to cycling infrastructure prioritisation in previous research, this is the first time that they have been combined, creating an integrated road segment prioritisation approach. The approach, which can be applied to other cities, as shown in the Appendix, is demonstrated in a case study of Manchester, resulting in cycling networks that balance directness against the need for safe and stress-free routes under different investment scenarios. A key benefit of the approach from a policy perspective is its ability to support egalitarian and cost-effective strategic cycle network planning. \\par\\textbf{Keywords:} cycling networks, low-traffic neighborhoods, routing, transport equity"
# keywords: "cycling, low-traffic neighborhoods, potential demand, routing"
output:
  # bookdown::github_document2:
  bookdown::pdf_document2:
  # # bookdown::word_document2:
      toc: false
      extra_dependencies: ["subcaption", "longtable", "float", "multirow", "placeins", "multicol"]
editor_options: 
  markdown: 
    wrap: 72
---

@@

The studies outlined above lay out the fundamentals for designing
cycling networks that generate significant cycling uptake, but they do
not propose network-level interventions. In this section we outline
methods used in past studies, namely optimization and network
analysis methods, such as connected components and community
detection, and examine how they are leveraged to suggest cycling network
designs. We compare the effectiveness of these network-level studies in
incorporating the fundamentals outlined above. 
Our proposed approach is inspired by these methods, but it attempts to add to them
by ensuring that all of the outlined fundamentals are accounted for. It
also goes further by attempting to factor in ethical considerations
relating to distribution of investment.

*Optimization* methods have been used to propose improvements to
cycling networks. @mesbah2012bilevel propose a bi-level formulation to
optimize allocation of cycling lanes to the network without exceeding a
set budget. They account for the effect of cycling lanes on car traffic,
and attempt to maximize utilization of said lanes with minimal impact on
car travel times. Safeguarding against increased car traffic may be counter-productive if the goal is to create a mode-shift, as research has shown that reducing road space for cars leads to less cars on the road, a phenomenon referred to as "traffic evaporation" [@nello2020environmental].
Car usage is not considered by @mauttone2017bicycle,
who develop an optimization framework that aims to minimize the total
user cost of cycling on the network. The aggregate flow^[*flow* is
used throughout this research to refer to the cycling demand when it is
routed onto the road network. The flow on any road segment is the
cumulative demand on it, resulting from cyclists commuting between
various OD pairs] on links^[*link*
refers to a road segment throughout this research] is obtained by using
shortest paths to route existing cycling demand onto the road network,
and the solution is a proposed set of links where cycling infrastructure
should be added in order to minimize the overall travel cost of cyclists
across the network. The cost of traversing a link is given as a function
of its length and whether or not it has cycling infrastructure. The
problem has also been solved by attempting to find the minimum cost of
improving roadway links to meet a desired level of service (LOS)
[@duthie2014optimization]. In this formulation, all origin-destination
(OD) pairs need to be connected by roads that meet the desired LOS. A
limitation of these problem formulations is that they do not explicitly solve for
continuity, which is dealt with using either a constraint specifying
that each link with a bike lane should be connected to at least one
destination [@mesbah2012bilevel], a constraint on maximum deviation
from shortest paths [@duthie2014optimization], or a discontinuity
penalty to prioritize connected road segments [@mauttone2017bicycle].

@@

Our work builds on these *community finding* approaches by proposing a
similar greedy network expansion algorithm for cycle network expansion
within communities. We incorporate community finding methods for study
area partitioning with weighted routing to avoid links that are
stressful to cycle on. In doing so, we propose an approach that
accounts for motivators and deterrents to cycling. We propose three
sub-methods that address on some of the limitations of previous
studies. These limitations include (a) bias inherent when basing network
design solely on existing cycling demand, (b) proposing routes that may
not correspond to studies on cyclist preference and government policies,
and (3) insufficient consideration of the ethical principles underlying
the analysis. Section \@ref(calculating-potential-cycling-demand)
focuses on calculating potential cycling demand. Section \@ref(routing)
focuses on routing the demand onto the road network while accounting for
cyclist preferences and government priorities. Section
\@ref(road-segment-prioritisation) outlines a method for partitioning the
study area based on a community finding algorithm and routed cycling
demand. It then introduces the network expansion algorithms, and compares an approach grounded in
'egalitarianism' to one grounded in 'utilitarianism'.

@@

\noindent where **d** and **s** are the distance and slope associated with each
OD pair and $\alpha$ and $\beta$ values are parameters calculated by a regression
model on training data. The square and square-root distance terms
"capture the non-linear impact of distance on the likelihood of
cycling", and interaction terms to capture the combined effect of slope
and distance [@lovelace2017propensity]. Alternative cycling uptake
models could be 'plugged in' to our approach for different contexts or scenarios of change.

@@

It should be reiterated that the weighting profile used for routing has
been developed for the purposes of this study. It creates a hierarchy of
road preference that is grounded in cyclist preferences and government
plans to create LTNs. A wide range of weighting profiles could be used
to represent different types of cyclists and road environments, as
described in @gehrke2020cycling and @furth2016network. Sensitivity
analysis could be done to determine an optimal weighting profile, but
given the variation in city road
networks<!-- (Figure \ref{fig:perc_person-km}) -->, this would probably
require calibration to the specific city. More accurate routing could be
carried out given the availability of road-level data. In such cases we
would add additional impedance to specific roads, giving more useful
routing results than the current approach which considers all roads
of the same type to be equivalent.

@@

Given that the "most essential activity entailed in the design of
cycle-friendly infrastructure is developing a cycle network"
[@parkin2018designing], we believe that the approach has great
potential to inform investment in cities such as Manchester where there
is political will to invest in cycling long-term. A benefit of the
approach is that it has relatively modest data requirements: only the
road network (from OSM), its topography (from satellite imagery), and
OD data (from the national census in this case) are
needed, opening up the possibility of deploying the approach in other
cities^[The results are easily reproducible for all UK cities, and can
also be reproduced for cities elsewhere given the availability of
commuter data. For instructions on reproducing the results shown in this
paper, see the **README** file in the R folder of the .zip file that
accompanies this paper (to be released as open source software when the
paper is published).].

@@

The approach can identify not only where there is high
potential for cycling but also trade-offs between
stress levels (related to motor traffic) and directness.
The results can therefore be used as a basis for recommendations on
road space reallocation *and* new infrastructure to unlock potential cycling demand.
The approach encourages consideration of a wider range of preferences
and needs than previous approaches that focus only on absolute potential.
Moreover, the inclusion of egalitarian principles in scenarios of change
encourages investment in cycling infrastructure to
increase the connectivity of existing cycling infrastructure
*and* investment that addresses geographical and social inequalities.
This ability to address inequalities in network prioritisation is particularly
important given research research showing substantial *inequalities*
around transport provision in general and cycling uptake and investment
in particular [@lucas2016method; @vidaltortosa_infrastructure_2020].

The approach is not without limitations. The level of detail is only as
good as the granularity of the available data (in this case relatively
coarse zones). @iacono2010measuring note that such large travel zones
are not ideal for understanding route choice behaviour of cyclists and
pedestrians. They also give rise to an \`ecological fallacy' whereby
average characteristics are assumed to apply to all residents of the
aggregated geographical area, suggesting a need for applying the methods
to more granular OD data (and for governments and other
data-collecting organisations to make OD data more
readily available).
The potential demand calculation is also based on cycling in the traditional sense, and does not consider the 
effect of micro-mobility on reducing topology-related impedance to cycling. Given that we are proposing an approach which can accommodate any cycling uptake functions, this is an acceptable limitation.
